---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
PerseusR provides an interface to use perseus data in R and vice versa

# Reading Perseus data in R

## As a list

The most basic way to import data from perseus into r is as a list whose 
elements can then be used in any sort of workflow.

```{r}
library(PerseusR)

dataFolder <- system.file('extdata', package = 'PerseusR')
dataFiles <- list.files(dataFolder, 
                        pattern = "matrix[[:digit:]]*.txt",
                        full.names=TRUE)
dataFile <- dataFiles[[1]]

default_output <- read.perseus.default(dataFile)
class(default_output)
print(default_output)
```

## As a MatrixData object

The custom `MatrixData` object is the most faithfull representation of a perseus
matrix in R and is an internal class used by this package to validate whether
the fields will be compatible with perseus in additional workflows.

```{r}
dm_out <- read.perseus.as.matrixData(dataFile)
class(dm_out)
print(dm_out)
```

The data matrix object contains the slots that are supported by perseus and can
be accessed by the methods of the object (not all of them must be present)

```{r}
main(dm_out)
annotRows(dm_out)
annotCols(dm_out)
PerseusR::description(dm_out) # Biobase has a descr function as well...
```


## As an expressionSet object (for bioconductor usage)

If you feel more confortable with the bioconductor expression set class, 
you can also import the data as such.

```{r}
require(Biobase)
eSet_out <- read.perseus.as.ExpressionSet(dataFile)
class(eSet_out)
print(eSet_out)

eSet_out@annotation
```

and the equivalentes with the perseus functions would go like this:

1. The `exprs` slot in the expressionSet objects would be equivalent to the `main` data frame in perseus.
2. The `featureData` would be equivalent to the `annotationCols`.
3. The `phenoData` would be aquivalent to the `annotationRows`.
4. The `Annotation` would be equivalent to the `descr`.

# Checking Data compatibility in R

There are a series of functions that check the compatibility R objects with
perseus.

## MatrixData 

This function would not provide much insight on this kind of object because
it is run when constructing it; Therefore if the data is not compatible it would 
return an error when trying to create it

```{r error=FALSE}

df <- matrixData(
    main = data.frame(a = 1:3, b = 6:8),
    annotCols = data.frame(c = c('a','b','c')),
    annotRows = data.frame(x = factor(c('1','1'))))

MatrixDataCheck(df)

my_error <- try({
  matrixData(
    main = data.frame(a = 1:3, b = 6:8, c = 1:3),
    annotCols = data.frame(c = c('a','b','c')),
    annotRows = data.frame(x = factor(c('1','1'))))
})

print(my_error)

```

## lists

In this case this functions would be usefull, since the object is not inherently 
compatible.


```{r, message=FALSE, warning=FALSE}
my_list <- list(main = data.frame(a = 1:3, b = 6:8),
    annotCols = data.frame(c = c('a','b','c')),
    annotRows = data.frame(x = factor(c('1','1'))))

MatrixDataCheck(my_list)

my_list <- list(main = data.frame(a = 1:3, b = 6:8, c = 1:3),
    annotCols = data.frame(c = c('a','b','c')),
    annotRows = data.frame(x = factor(c('1','1'))))

my_error <- try({
  MatrixDataCheck(my_list)
})
print(my_error)

```

## expressionSet objects

Single matrix Expression Sets will usually be compatible, since most of the 
restriction in  matrix data objects also apply to expression set objects
(regarding the dimensions and classes of the objects in each of the slots) 

```{r}
eSet <- Biobase::ExpressionSet(matrix(1:10, ncol = 2))

MatrixDataCheck(eSet)
```



# Writting Data into Perseus-compatible text representations

This is the way in which perseus will read the data again so here are a couple
of examples.

## MatrixData

```{r}
# Here you can use any tipe of conection, similar to th base write... functions
# The usage should be fairly similar to the write.table function
# 

write.perseus(dm_out, 'my_output_df.txt')

# which would output somethint like this

cat(readLines('my_output_df.txt'), sep = '\n')
```

## data.frame

Data frames are converted in such manner that numeric columns are
transfered as the main DF and non numerics as the annotation cols
of the perseus DF

```{r}

my_df <- data.frame(Con1 = 1:3, Con2 = 4:6, An1 = letters[1:3])

write.perseus(my_df, con = 'my_output_df.txt')

cat(readLines('my_output_df.txt'), sep = '\n')
```

## matrix

Numeric matrices can be outputed as well :D

```{r}
my_matrix <- matrix(1:10, ncol = 2, 
                    dimnames = list(letters[11:15], letters[1:2]))

write.perseus(my_matrix, con = 'my_output_df.txt')

cat(readLines('my_output_df.txt'), sep = '\n')
```

## list

Lists are a little trickier, currently it looks for named elements that
match the arguments, so it would use the elements named: `main` `annotCols`
`annotRows` and `descr`

```{r}

my_list <- list(main = data.frame(A = 1:5, B = 6:10),
                annotRows = data.frame(is_control = c(TRUE, FALSE)),
                annotCols = data.frame(Names = letters[1:5]),
                descr = c('something',
                          'something else',
                          'yet another thing'))

write.perseus(my_list, con = 'my_output_df.txt')
cat(readLines('my_output_df.txt'), sep = '\n')


```

## expressionSet objects

```{r}
eSet <- Biobase::ExpressionSet(matrix(1:10, ncol = 2))

write.perseus(eSet, con = 'my_output_df.txt')
cat(readLines('my_output_df.txt'), sep = '\n')

```

## Adding elements to the output

Since not all data typer support allthe elements, one can specify them to the
function and they will be passed over to the output.

```{r}
my_matrix <- matrix(1:10, ncol = 2, 
                    dimnames = list(letters[11:15], letters[1:2]))
my_annotations_rows <- data.frame(My_Names = letters[1:2])
my_annotations_cols <- data.frame(My_Genes = letters[11:15])

write.perseus(my_matrix, con = 'my_output_df.txt', 
              annotCols = my_annotations_cols,
              annotRows = my_annotations_rows)

cat(readLines('my_output_df.txt'), sep = '\n')
```

## Removing elements form the output

Elements can be set to `NULL` to remove that section from the output

```{r}
my_matrix <- matrix(1:10, ncol = 2, 
                    dimnames = list(letters[11:15], letters[1:2]))

write.perseus(my_matrix, con = 'my_output_df.txt', 
              annotCols = NULL)

cat(readLines('my_output_df.txt'), sep = '\n')
```


```{r, message=FALSE, warning=FALSE, include=FALSE}
try({file.remove('my_output_df.txt')})
```
